---
title: "Data Analysis Big Questions"
author: "Julia Lerner"
date: "11/03/2019"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.


```{r}
# installing and loading packages that i might need
#install.packages("tidycensus")
#install.packages("rgdal")
#install.packages("rgdal")
#install.packages("tidyverse")
#install.packages("arcos")
#install.packages("scales")
#install.packages("ggrepel")
#install.packages("mapview")
#install.packages("ggthemes")
library(tidyverse)
library(janitor)
library(arcos)
library(scales)
library(ggrepel)
library(tidycensus)
library(mapview)
library(ggthemes)

```

```{r}
# store one of our API keys as an object called key
key <- "uO4EK6I"

arcos_county_pills_per_year <- summarized_county_annual(key = key) %>%
  clean_names()

arcos_buyer_yearly <- combined_buyer_annual(key=key) %>%
  clean_names()

```

```{r}
# load data fr buyer address, buyer totals, buyer annual by year, and state population per year
buyer_address <- read_csv("data/buyer_addresses.csv")
buyer_annual_by_year <- read_csv("data/buyer_annual_by_year.csv")
buyer_totals <- read_csv("data/buyer_totals.csv")
state_population_per_year <- read_csv("data/state_population_per_year.csv")
babies_per <- read_csv("data/delivery_hospitalizations.csv")
children_in_foster_care <- read_csv("data/children_in_foster_care.csv")
state_population <- read_csv("data/state_population.csv")
state_names <- read_csv("data/50_us_states_all_data.csv")
```


```{r}
# How many children (especially in states with high-opioid addiction rates) have gone into foster care in the years 2006-2012? 
# this is for nationwide, how many childen are in foster care for 2006-2012. 
children_in_foster <- children_in_foster_care %>%
  filter(TimeFrame %in% c("2006", "2007", "2008", "2009", "2010", "2011", "2012") & Location=="United States")

# now lets focus in on a particular state. According to the National Inpatient Sample and State Inpatient Database, Healthcare Cost and Utilization Project info, Vermont has the highest rate of change in babies per 1000 born addicted to opioids. Let's start there. 
```

```{r}
#How many children are in Vermont's foster system in the years 2006-2012?
children_in_VTfoster_2006 <- children_in_foster_care %>%
  filter(TimeFrame %in% c("2006", "2007", "2008", "2009", "2010", "2011", "2012") & Location=="Vermont") %>%
  rename(year = TimeFrame) %>%
  rename(total_kids = Data)
```

```{r}
# and how many pills were sent to vermont in this time frame? 
buyer_annual_year_VT <- buyer_annual_by_year %>%
  rename(State = buyer_state) %>%
  filter(State =="VT") %>%
  mutate(State = "Vermont") %>%
  group_by(year, State) %>%
  summarise(shipments = n(),
            total_pills = sum(dosage_unit)) 
```

```{r}
state_population_per_year_UPDATE <- state_population_per_year %>%
 rename(state = buyer_state)
```

```{r}

# the highest number of pills by year ws in 2009. was that also the year with the highest number of children in the system? 

kids_and_pills <- buyer_annual_year_VT %>%
  inner_join(children_in_VTfoster_2006, by="year") %>%
  select(year, total_pills, Location, total_kids) %>%
  arrange(desc(year)) %>%
  rename(state = Location)

# interestingly the highest population of children in foster care was in 2006, when there were 1,379 kids in foster care. In 2009, when 20,783,165 pills flooded VT, there were only 1062 kids in foster care. 

```

```{r}
# okay but how percent of people in vermont were in foster care in this time frame?

#this is the data
state_population_per_year_UPDATE <- state_population_per_year_UPDATE %>%
  rename(Location = state) %>%
  filter(Location=="VT") %>%
  mutate(Location="Vermont") 

kids_and_pop <- children_in_VTfoster_2006 %>%
  full_join(state_population_per_year_UPDATE, by ="Location") %>%
  filter(Location=="Vermont") %>%
  mutate(total_kids = as.numeric(total_kids)) %>%
  mutate(rate = (total_kids/population_average)*10000)

# this is the chart
ggplot(kids_and_pop) +
  geom_bar(stat="identity", aes(year, rate), fill="navy") +
  labs(x="year", y="rate of foster children per 10,000 in Vermont", title="In Vermont, the number of foster children per 10,000 in Vermont fell starting in 2006", subtitle = "After 2010, the number of foster children began to rise", caption = "Source: Census data") +
  scale_x_continuous(breaks = c(2006, 2007, 2008, 2009, 2010, 2011, 2012)) +
  scale_y_continuous(labels = comma)
```

```{r}
#how many pills went to vermont in 2006-2012? how many pills per person were sent to vermont in this time frame? 
buyer_annual_year_VT <- buyer_annual_year_VT %>%
  clean_names() %>%
  glimpse


state_population_per_year_UPDATE <- state_population_per_year_UPDATE%>%
  rename(state=Location) %>%
  glimpse


buyer_totals_vt <- buyer_annual_year_VT %>%
 inner_join(state_population_per_year_UPDATE, by="state") %>%
 group_by(state, population_average) %>%
 mutate(rate = total_pills/population_average) %>%
 arrange(desc(rate)) %>%
  filter(state =="Vermont")


# total number of pills: 123414703
# rate per person = 197.8865 TOTAL
#rate per year varies
 
# I'd like to make a chart showing the relationship between population and pills in the state. Let's make a bar graph of total population per year and then a line that shows the rate of pills per year 

ggplot(buyer_totals_vt) +
  geom_bar(stat="identity", aes(year, rate), fill="black") +
  labs(x="year", y="pills per person", title="In Vermont, the rate of pills per person peaked in 2009", subtitle = "In 2010, the number of pills per person began to fall", caption = "Source: Census data") +
  scale_x_continuous(breaks = c(2006, 2007, 2008, 2009, 2010, 2011, 2012)) +
  scale_y_continuous(labels = comma)

```

```{r}
# this is the data to load the map below 
# Define API Key
census_api_key("549950d36c22ff16455fe196bbbd01d63cfbe6cf")

# If you need to look up variables, this is how you do it
acs_variables <- load_variables(2017, "acs5" )

county_geodata <- get_acs(geography = "county",
              variables = "B01001_001", geometry = TRUE)

state_geodata <- get_acs(geography = "state",
              variables = "B01001_001", geometry = TRUE)

```

```{r}
children_in_foster_care_CA <- children_in_foster_care %>%
  filter(Location=="California") %>%
  filter(TimeFrame>="2006") %>%
  filter(TimeFrame<="2012")
```
```{r}
# what's next? Let's look at children in foster care nationally

# turn off scientific notation 
options(scipen=999)
# this is the sheet that the map will pull the data from. 
children_in_foster_care_STATES <- children_in_foster_care %>%
  filter(LocationType =="State") %>%
  rename(total_kids = Data) %>%
  rename(year=TimeFrame) %>%
  rename(state = Location) %>%
  mutate(total_kids=as.numeric(total_kids))

#load the state geometry
state_geodata <- get_acs(geography = "state",
    variables = "B01001_001", geometry = TRUE, shift_geo = TRUE) %>%
  rename(state = NAME) 

# now let's do a full join of foster care data and state shapes

state_foster_pops <- state_geodata %>% 
  full_join(children_in_foster_care_STATES, by="state") %>%
filter(year=="2006")

# this is the map
state_foster_pops %>%
  ggplot(aes(fill = total_kids)) +
  geom_sf(lwd = 0) +
  theme_map() +
  labs(fill='Population',title="California, Texas have largest populations of children in foster care", subtitle = "2017 population, U.S. Census", caption = "Source: U.S. Census ACS") +
   theme(legend.position="none")

```


```{r}
# how many counties in the U.S. received more than 250,000 pills between 2006 and 2012? 

arcos_county_pills_per_year <- summarized_county_annual(key = key)

US_pharmacies_over250k <- arcos_county_pills_per_year %>%
  select(buyer_state, buyer_county, dosage_unit, year) %>%
  arrange(desc(dosage_unit)) %>%
  filter(dosage_unit > 250000)

# there are 17,000 counties that received more than that. what if we look at which states had the most foster kids to decide which states to look at?

children_in_foster_care_STATES %>%
  mutate(total_kids = as.numeric(total_kids)) %>%
  filter(year=="2006") %>%
  arrange(desc(total_kids))

# it looks like California, Texas, NY, Florida, Pennsylvania, Michigan, Illinois, and Ohio had the highest rates of children in foster care.  I want to look at California counties now to see if the rate of pills per foster kid is higher in california than in other states. 

US_pharmacies_CA <- arcos_county_pills_per_year %>%
  select(buyer_state, buyer_county, dosage_unit, year) %>%
  arrange(desc(dosage_unit)) %>%
  filter(buyer_state=="CA", year=="2006")

# now, going back to VT for similar analysis- which counties received the most pills in Vermont? 

vt_pharmacies_over250k <- arcos_county_pills_per_year %>%
  select(buyer_state, buyer_county, dosage_unit, year) %>%
  arrange(desc(dosage_unit)) %>%
  filter(dosage_unit > 250000) %>%
  filter(buyer_state=="VT")




```

```{r}

# lets find the rate of pills per foster kid in california in 2006.
foster_per_year_CA <- children_in_foster_care_STATES %>%
  filter(state=="California") %>%
   mutate(total_kids = as.numeric(total_kids)) %>%
  glimpse

buyer_yearly_CA <- arcos_buyer_yearly %>%
  filter(buyer_state=="CA") %>%
  mutate(buyer_state="California") %>%
  group_by(year, buyer_state) %>%
  summarise(shipments = n(),
            total_pills = sum(dosage_unit)) %>%
  rename(state=buyer_state) %>%
  glimpse

pills_foster_CA <- buyer_yearly_CA %>%
 inner_join(foster_per_year_CA, by="year") %>%
 #group_by(state, total_kids) %>%
 mutate(rate = total_pills/total_kids) %>%
 arrange(desc(rate)) 

# so in California in 2006, there were 11,965.49 pills per foster kid. That seems really high! What does that same number look like in Vermont?

```

```{r}
# running same questions as above, but for Vermont instead of California. I want a point of comparison. 
foster_per_year_VT <- children_in_foster_care_STATES %>%
  filter(state=="Vermont") %>%
   mutate(total_kids = as.numeric(total_kids)) 

buyer_yearly_VT <- arcos_buyer_yearly %>%
  filter(buyer_state=="VT") %>%
  mutate(buyer_state="Vermont") %>%
  group_by(year, buyer_state) %>%
  summarise(shipments = n(),
            total_pills = sum(dosage_unit)) %>%
  rename(state=buyer_state) 

pills_foster_VT <- buyer_yearly_VT %>%
 inner_join(foster_per_year_VT, by="year") %>%
 mutate(rate = total_pills/total_kids) %>%
 arrange(desc(rate)) 

# so in the same year, there were 10,674.07 pills per foster kid sent to Vermont. These numbers are pretty similar, which is kind of shocking when you consider that there were only 1379 foster kids in Vermont in 2006, and 78373 foster kids in California in 2006. 

# We really seem to mostly be focusing on the rust belt for opioid data. I'm thinking someone should look at VT, especially given the context of numbers when compared to California. 
```


```{r}
#join pills_per_person with foster_rate_pop_VT
state_population_per_year <- state_population_per_year %>%
 rename(state=buyer_state)

pills_per_person <- state_population_per_year %>%
  full_join(buyer_annual_year_VT, by="state") %>%
  mutate(rate=total_pills/population_average) %>%
  filter(state=="Vermont") %>% glimpse

# set up state population per year
state_population_per_year <- state_population_per_year %>%
  #rename(state=buyer_state) %>%
  filter(state=="VT") %>%
  mutate(state="Vermont") 

#set up population for fosters per year. 
foster_rate_pop_VT <- children_in_foster_care_STATES %>%
  full_join(state_population, by="state") %>%
  filter(state=="Vermont") %>%
  mutate(rate=(total_kids/population_average)*1000) %>%
  filter(year>="2006") %>%
  filter(year<="2012") 

# ok now that you have these two data sets, you need to do a scatterplot with them. first, inner join. then make the plot 

scatter_pills_kids <- pills_per_person %>%
  inner_join(foster_rate_pop_VT, by ="year")  %>%
  select(state.x, year, population_average.x, total_pills, total_kids, rate.x, rate.y) %>%
  rename(pills_per_person=rate.x) %>%
  rename(foster_per_person=rate.y)

```

```{r}
# here's the scatterplot for the foster rate and pills per person in VT
ggplot(scatter_pills_kids) +
  geom_point(aes(foster_per_person, pills_per_person)) +
  geom_smooth(aes(foster_per_person, pills_per_person), method = "lm", se = FALSE)  +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = comma)  +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_text_repel(aes(foster_per_person, pills_per_person, label=foster_per_person),
                  subset(scatter_pills_kids)) +
  labs(x="fosters per 1000 people in VT", y="pills per person", title="as opioid use decreases, so do number of kids in foster care", caption = "Source: DEA ARCOS database, via Washington Post")
```

```{r}
# it definitely looks like there is a relationship there. 
```

```{r}
# so we're gonna move on and try to figure out some more stuff. Let's go back to vermont. Let's find the rate of pills per person per county. 

VT_counties <- arcos_county_pills_per_year %>%
  filter(buyer_state=="VT") %>%
  rename(state=buyer_state) %>%
  rename(county=buyer_county)

Vermont_pills_rate <- VT_counties %>%
  inner_join(state_population_per_year_UPDATE, by="state") %>%
  mutate(rate=(dosage_unit/population_average))

#are there any counties or years that stand out?

Vermont_pills_rate_2006 <- Vermont_pills_rate %>%
  filter(year=="2006")
# chitenden county seems a lot higher than other counties

Vermont_pills_rate_2007 <- Vermont_pills_rate %>%
  filter(year=="2007")

Vermont_pills_rate_2008 <- Vermont_pills_rate %>%
  filter(year=="2008")

Vermont_pills_rate_2009 <- Vermont_pills_rate %>%
  filter(year=="2009")

Vermont_pills_rate_2010 <- Vermont_pills_rate %>%
  filter(year=="2010")

Vermont_pills_rate_2011 <- Vermont_pills_rate %>%
  filter(year=="2011")

Vermont_pills_rate_2012 <- Vermont_pills_rate %>%
  filter(year=="2012")
  

# so now lets look at chittenden 
Vermont_pills_rate_Chittenden <- Vermont_pills_rate %>%
  filter(county=="CHITTENDEN") %>%
  glimpse

# lets also look at Essex, which seemed to receive the least. 
Vermont_pills_rate_Essex <- Vermont_pills_rate %>%
  filter(county=="ESSEX")

# is there a particular year in Chittenden that stood out? 
#In 2009, the rate of pills per person almost doubled, from 7.2/person in 2008 and 12.6/person in 2009. 
# now lets do some reporting to find out why. 
  
```

```{r}
# important numbers

babies_per %>% glimpse
# The number of babies per 1k born addicted to opioids in VT
# in 2006: 12.9
# in 2007: 14.6
# in 2008: 19
# in 2009: 28.5
# in 2010: 27.1
# in 2011: 33.8
# in 2012: 43.7

# The number of babies per 1k born addicted to opioids in CA:
# in 2006: 1.1
# in 2007: 1.0
# in 2008: 1.1
# in 2009: 1.2
# in 2010: 1.3
# in 2011: 1.6
# in 2012: n/a

children_in_foster_care %>% glimpse
#Nationwide children in foster care:
# in 2006: 510885
# in 2007: 486619
# in 2008: 459828
# in 2009: 423773
# in 2010: 408425
# in 2011: 397885
# in 2012: 397091

#Children in foster care in Vermont: 
# in 2006: 1379
# in 2007: 1309
# in 2008: 933
# in 2009: 1062
# in 2010: 1200
# in 2011: 1010
# in 2012: 975

#Rate of children in foster care in Vermont per 10000 people: 
# in 2006: 22.11
# in 2007: 20.99
# in 2008: 14.96
# in 2009: 17.03
# in 2010: 19.24
# in 2011: 16.19
# in 2012: 15.63

pills_foster_VT %>% glimpse
# number of pills sent to vermont: 
# in 2006: 14719540
# in 2007: 16493890
# in 2008: 17106400
# in 2009: 20783165
# in 2010: 18037408
# in 2011: 18383830
# in 2012: 17890470

#rate of pills per foster kid in Vermont: 
# in 2006: 10674.07
# in 2007: 12600.37
# in 2008: 14255.33
# in 2009: 19569.84
# in 2010: 19332.70
# in 2011: 18201.81
# in 2012: 18349.20

# rate of pills per person in vermont: 
# in 2006: 23.60
# in 2007: 26.45
# in 2008: 27.43
# in 2009: 33.32
# in 2010: 28.92
# in 2011: 29.45
# in 2012: 28.69

pills_foster_CA %>% glimpse
# number of pills sent to California: 
# in 2006: 1318766136
# in 2007: 1301822983
# in 2008: 1198384341
# in 2009: 1167527178
# in 2010: 1095626025
# in 2011: 1037393654
# in 2012: 937771106

# rate of pills per foster kid in California: 
# in 2006: 24292.04
# in 2007: 23775.85
# in 2008: 20766.35
# in 2009: 19394.78
# in 2010: 16274.17
# in 2011: 14019.21
# in 2012: 11965.49

children_in_foster_care_CA %>% glimpse
#number of children in California foster system
# in 2006: 78373
# in 2007: 73998
# in 2008: 67323
# in 2009: 60198
# in 2010: 57708
# in 2011: 54754
# in 2012: 54288


```






